Galois Slicing as Differentiation

### 2025-07-11 main/language.tex
[ tex ] more tweaking of the examples

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes:
  \begin{displaymath}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \{\top,\bot\} \multimap \mathbb{1} \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times \mathbb{1}
  \end{displaymath}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input $\mathsf{a}$ with the top approximation maps indeed reveals that
  only the numbers in the rows tagged with that label in the input are used, and likewise for $\mathsf{b}$.
\item More quantitative approximation information can be obtained by using the interval approximation
  interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}.

  Recall that $\mathrm{query}\,\mathsf{a}\,\mathit{db} = 1$, so in the reverse direction we can pick an
  interval containing $1$ to discover the largest (i.e. least in the order) intervals that will give rise to
  this output as \emph{independent} changes to the input. For example, if we pick
  $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Indicating that, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$ for this query, either the first
  $\mathsf{a}$ row could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in
  $[\frac{9}{10},\frac{11}{10}]$, and it does not matter what the number in the second entry was.
\end{enumerate}

We have tested these examples on our fully formalised Agda implementation. See the file \texttt{example.agda}
in the supplementary material for details.

>>>>

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes:
  \begin{displaymath}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \{\top,\bot\} \multimap \mathbb{1} \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times \mathbb{1}
  \end{displaymath}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input ``$\mathsf{a}$'' at approximation $\top$ reveals that only the
  numbers in the rows tagged with ``$\mathsf{a}$'' in the input are used, and likewise for ``$\mathsf{b}$''.
\item Quantitative approximation information with non-trivial dependency can be obtained by using the interval
  approximation interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}. Recall that $\mathrm{query}\,(\mathsf{a},\mathit{db}) = 1$, so in the reverse
  direction we must choose an interval containing $1$ to discover the largest (i.e. least in the order)
  intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we
  pick $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Thus, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$, either the first ``$\mathsf{a}$'' row
  could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in $[\frac{9}{10},\frac{11}{10}]$, and
  the number in the ``$\mathsf{b}$'' row is not relevant.
\end{enumerate}

We have tested these examples on our Agda implementation. See the file \texttt{example.agda}.
